Simulating High-Dimensional Multivariate Data

### 2021-03-01 010-background.Rmd
fixed formatting

```{r ch010-realDataFig, cache=FALSE, fig.width=8, fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
ggpairs(
  data = example_brca[, gene_sample],
  upper = list(continuous = wrap('cor', method = "spearman"))
) + theme_bw()
```

@@

Spearman's correlation for discrete marginal suffers issues whenever there is a nonzero probability of ties. For example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1 [@MB13]. One remedy for this issue is to rescale the Spearman correlation. For two random variables $X,Y$ with probability mass functions (PMFs) or probability densities functions (PDFs) $p(x)$ and $q(y)$, respectively, define the rescaled Spearman's correlation as

@@

Note that with continuous marginals the rescaling returns $\rho_S$. For discrete marginals with large or infinite support, computing the adjustment factors $\sum_x p(x)^3$, $\sum_y p(y)^3$ over all large number of pairs becomes expensive (often violating desired property SP1). Further, the infinite sums must be approximated for count-valued data.

>>>>

```{r ch010-realDataFig, cache=FALSE, fig.width=8, fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
ggpairs(
  data = example_brca[, gene_sample],
  upper = list(continuous = wrap('cor', method = "spearman"))
) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5))
```

@@

Spearman's correlation for discrete marginal distributions suffers issues whenever there is a nonzero probability of ties. For example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1 [@MB13]. One remedy for this issue is to rescale the Spearman correlation. For two random variables $X,Y$ with probability mass functions (PMFs) or probability densities functions (PDFs) $p(x)$ and $q(y)$, respectively, define the rescaled Spearman's correlation as

@@

Note that with continuous marginals the rescaling returns $\rho_S$. For discrete marginals with large or infinite support, computing the adjustment factors $\sum_x p(x)^3$, $\sum_y p(y)^3$ over a large number of pairs becomes expensive (often violating desired property SP1). Further, the infinite sums must be approximated for count-valued data.

### 2021-03-01 030-bigsimr-package.Rmd
fix comparison with theory in 030
fixed formatting

First we use `cor_bounds` to ensure that the pairwise target correlation values are valid prior to mapping step which constructs $R_X$ for the MVN input into NORTA (see Section \@ref(algorithms)). `cor_bounds` estimates the pairwise lower and upper correlation bounds using the Generate, Sort, and Correlate algorithm of @DH2011.

>>>>

First we use `cor_bounds` to ensure that the pairwise target correlation values are valid prior to the mapping step which constructs $R_X$ for the MVN input into NORTA (see Section \@ref(algorithms)). `cor_bounds` estimates the pairwise lower and upper correlation bounds using the Generate, Sort, and Correlate algorithm of @DH2011.

### 2021-03-01 040-monte-carlo-evaluation.Rmd
fixed formatting

```{r ch040-LoadLib040, include=FALSE}
box::use(
    dplyr[...],
    ggplot2[...],
    patchwork[...],
    tidyr[pivot_longer],
    bigsimr[bigsimr_setup, distributions_setup],
    knitr[kable],
    RColorBrewer[brewer.pal]
)
load("data/bivariate_normal_sims.rda")
load("data/bivariate_gamma_sims.rda")
load("data/bivariate_nbinom_sims.rda")
load("data/benchmark_dependences.rda")

@@

kable(tabMAE, booktabs = TRUE,
      col.names = c("No. of random vectors",
                    "Correlation type",
                    "Distribution",
                    "Mean abs. error"),
      caption = "Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.")
```

Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation values all three dependency measures, at least in these limited simulation settings for the rank-based correlations. Our Pearson matching performs nearly as well as Spearman or Kendall, except a slight increase in error for negative binomial case. This is due the particularly large counts generated with our choice of parameters for $p$ and $r$.

@@

<!-- 
See [Discussion](discussion) for future directions for fast Pearson matching and discrete-specific modifications.
-->

@@

With information of our method's accuracy from a low-dimensional perspective, we now turn to assessing whether the `bigsimr` can scale to larger dimensional problems with practical computation times. Specifically, we ultimately generate $B=1,000$ random vectors for $d=\{100, 250, 500, 1000, 2500, 5000, 10000\}$ for each correlation type, $\{Pearson, Spearman, Kendall\}$ while timing the algorithm's major steps.

@@

tabTimes %>%
    kable(booktabs = TRUE,
          col.names = c("Dimension",
                        "Correlation type",
                        "Total Time (Seconds)"),
          caption = 'Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
```

The results larger dimensional vectors show scalability to ultra-high dimensions for all three correlation types, although the total times do become much larger. 
Figure \@ref(fig:ch040-largeDfig) displays computation times for $d=\{1000, 2500, 5000, 10000\}$. For $d$ equal to 1000 and 2500, the total time is under a couple of minutes. At $d$ of 5000 and 10,000, Pearson correlation matching in the *Adjust Correlation* step becomes costly. Interestingly, Pearson is actually faster than Kendall for $d=10,000$ due to bottlenecks in *Compute Correlation* and *Check Admissibility*. Uniformly, matching Spearman correlations is faster, with total times under 5 minutes for $d=10,000$, making Spearman correlations the most computationally-friendly dependency type. With this in mind, we scaled the simulation to $d=20,000$ for the Spearman type and obtained the 1,000 vectors in under an hour (data not shown). In principle, this would enable the simulation of an entire human-derived RNA-seq data set. We note that for a given target correlation matrix and margins, steps i, 1, and 2 only need to be computed once and the third step, *Simulate Data*, is fast under all schemes for all $d$ under consideration.

>>>>

```{r ch040-LoadLib040, include=FALSE, cache=FALSE}
box::use(
    dplyr[...],
    ggplot2[...],
    patchwork[...],
    tidyr[pivot_longer],
    bigsimr[bigsimr_setup, distributions_setup],
    knitr[kable],
    RColorBrewer[brewer.pal]
)
load("data/bivariate_normal_sims.rda")
load("data/bivariate_gamma_sims.rda")
load("data/bivariate_nbinom_sims.rda")
load("data/benchmark_dependences.rda")

@@

kable(tabMAE, booktabs = TRUE,
      linesep = c("", "", "\\addlinespace"),
      col.names = c("No. of random vectors",
                    "Correlation type",
                    "Distribution",
                    "Mean abs. error"),
      caption = "Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.")
```

Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation values all three dependency measures, at least in these limited simulation settings for the rank-based correlations. Our Pearson matching performs nearly as well as Spearman or Kendall, except for a slight increase in error for negative binomial case. This is due the particularly large counts generated with our choice of parameters for $p$ and $r$.

@@

@@

With information of our method's accuracy from a low-dimensional perspective, we now turn to assessing whether `bigsimr` can scale to larger dimensional problems with practical computation times. Specifically, we ultimately generate $B=1,000$ random vectors for $d=\{100, 250, 500, 1000, 2500, 5000, 10000\}$ for each correlation type, $\{Pearson, Spearman, Kendall\}$ while timing the algorithm's major steps.

@@

tabTimes %>%
    kable(booktabs = TRUE,
          linesep = c("", "", "\\addlinespace"),
          col.names = c("Dimension",
                        "Correlation type",
                        "Total Time (Seconds)"),
          caption = 'Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
```

The results with larger dimensional vectors show scalability to ultra-high dimensions for all three correlation types, although the total times do become much larger. 
Figure \@ref(fig:ch040-largeDfig) displays computation times for $d=\{1000, 2500, 5000, 10000\}$. For $d$ equal to 1000 and 2500, the total time is under a couple of minutes. At $d$ of 5000 and 10,000, Pearson correlation matching in the *Adjust Correlation* step becomes costly. Interestingly, Pearson is actually faster than Kendall for $d=10,000$ due to bottlenecks in *Compute Correlation* and *Check Admissibility*. Uniformly, matching Spearman correlations is faster, with total times under 5 minutes for $d=10,000$, making Spearman correlations the most computationally-friendly dependency type. With this in mind, we scaled the simulation to $d=20,000$ for the Spearman type and obtained the 1,000 vectors in under an hour (data not shown). In principle, this would enable the simulation of an entire human-derived RNA-seq data set. We note that for a given target correlation matrix and margins, steps i, 1, and 2 only need to be computed once and the third step, *Simulate Data*, is fast under all schemes for all $d$ under consideration.
